**Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data**

### 2019-07-19 submission/manuscript.Rmd
update paper for table

In this study we established a rigorous ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome. We built on others' work in the microbiome field to set-up standards for developing and evaluating ML models for microbiome data [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019]. First, we used a held-out test set to illustrate the difference between cross-validation and testing AUROC values. When the difference between cross-validation and test performance is low, this suggest the models are not overfit and that they will perform similar with similar data. In all seven models, the difference between median cross-validation and testing AUROC values did not exceed `r paste(round(abs(difference_model[difference_index[7],1]), 3))` which suggests that these models are generalizable and can be used to test similar new data. Second, we performed the initial 80%-20% random datasplit 100 times in our ML pipeline. The randomization of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop robust ML models and to report reliable performance metrics. Depending on how the data is split, there is the chance of being overoptimistic about the predictive performance of a model. In our study, we showed that there was variability in AUROC values between different random data-splits in each of the models we tested. Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average between different data-splits. Third, we used the AUROC metric in our study to evaluate the predictive performance of the ML models. AUROC is always random at the value 0.5 and is a robust metric when a dataset is imbalanced. We also performed a full grid search for hyperparameter settings when building a ML model. Default hyperparameter settings in previously developed ML packages in R, Python, and Matlab programming languages are inadequate for effective application of classification algorithms and need to be optimized for each new dataset used to generate a model. In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default.

In this study, we benchmarked seven ML models with different classification algorithms to show that we should use ML models based on the goal of the study and our expectations of predictive performance, interpretability and computational burden. Microbiome studies use ML models with a classification task to learn the training data to assign labels, such as healthy or not to new data, but also to learn which features are important to discriminate between labels [@zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018]. Our results show that if the goal of a study is to learn the ecology behind a disease and to identify microbial biomarkers, we could create ML models that are inherently interpretable without losing predictive power. In terms of predictive performance, random forest model had the best testing AUROC values compared to the other 6 models. However, the second best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest. In terms of interpretability, random forest was a complex ML model and could only be explained using methods such as permutation importance. On the other hand, L2-regularized logistic regression was easier to interpret (i.e. ranking absolute regression coefficients of the trained model).

Even with interpretable models such as L2-regularized logistic regression, we need to be careful with how we treat the information we get from the models. In this study we used two different methods to interpret our linear models; ranking each OTU by (1) their absolute weights in the trained models and (2) their impact on the predictive performance based on permutation importance. We observed differences in the OTU rankings between the two interpretation methods due to collinearity in the dataset. Collinearity in a microbial dataset is when one OTU is highly correlated with another OTU which causes feature weights to not be unique. The feature weights of correlated OTUs are influenced by one another which makes it difficult to interpret the ML model. To avoid misinterpreting the models we should use the highly ranked correlated OTUs to generate hypotheses about the ecology of the disease and test them with follow-up experiments.

There are other criteria when choosing ML models such as the computational burden of developing it and the sample size. In terms of computational burden, random forest model trained each data-split in 83.2 hours whereas L2-regularized logistic regression trained in 12 minutes. The generalization performance of ML models depends on sample size. The more complex the model, the more data it will need. The dataset we used for our study had 490 samples, however microbiome studies that have smaller sample sizes would benefit from using less complex models such as L2-regularized logistic regression.

>>>>

Microbiome studies use ML models, often with a classification task to predict a phenotype but also to learn which microbes are important to discriminate that phenotype from others [@zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018]. Achieving either of these tasks in the context of a disease have far-reaching impact on human health, however ML as a tool in microbiome studies is still at its infancy. There is a need for a framework for developing rigorous ML models, identifying and overcoming potential pitfalls. Previous studies generated robust workflows to allow ML to be widely used by the microbiome researchers [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019] and in this study we set-up standards for developing and evaluating ML models for microbiome data [Table 1].

We benchmarked seven ML models with different classification algorithms to show that the goal of the microbiome study and the clearly defined ML problem should inform our model selection. Our results showed that if the goal of a study is to learn the ecology behind a disease and to identify microbial biomarkers, we could create ML models that are inherently interpretable without losing predictive power. In terms of predictive performance, random forest model had the best testing AUROC values compared to the other 6 models. However, the second best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest. In terms of interpretability, random forest was a complex ML model and could only be explained using methods such as permutation importance. On the other hand, L2-regularized logistic regression was easier to interpret by ranking absolute regression coefficients of the trained model.

Even with interpretable models such as L2-regularized logistic regression, we need to be careful with how we treat the information we get from the models. If we have a priori hypotheses based on knowledge from domain experts and we know the confounding variables in our dataset, we should incorporate that to our model interpretation. In this study we used two different methods to interpret our linear models; ranking each OTU by (1) their absolute weights in the trained models and (2) their impact on the predictive performance based on permutation importance. We observed differences in the OTU rankings between the two interpretation methods due to collinearity in the dataset. Collinearity in a microbial dataset is when one OTU is highly correlated with another OTU which causes feature weights to not be unique. The feature weights of correlated OTUs are influenced by one another which makes it difficult to interpret the ML model. To avoid misinterpreting the models we should use the highly ranked correlated OTUs to generate hypotheses about the ecology of the disease and test them with follow-up experiments.

Another criteria when choosing the ML models was the computational time it would take to develop it. Random forest model trained each data-split in 83.2 hours whereas L2-regularized logistic regression trained in 12 minutes.

We established a rigorous ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome. We used a held-out test set to illustrate the difference between cross-validation and testing AUROC values. When the difference between cross-validation and test performance is low, this suggest the models are not overfit and that they will perform similar with similar data. In all seven models, the difference between median cross-validation and testing AUROC values did not exceed `r paste(round(abs(difference_model[difference_index[7],1]), 3))` which suggests that these models are generalizable and can be used to test similar new data. Second, we performed the initial 80%-20% random datasplit 100 times in our ML pipeline. The randomization of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop robust ML models and to report reliable performance metrics. Depending on how the data is split, there is the chance of being overoptimistic about the predictive performance of a model. In our study, we showed that there was variability in AUROC values between different random data-splits in each of the models we tested. Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average between different data-splits. Third, we used the AUROC metric in our study to evaluate the predictive performance of the ML models. AUROC is always random at the value 0.5 and is a robust metric when a dataset is imbalanced. We also performed a full grid search for hyperparameter settings when building a ML model. Default hyperparameter settings in previously developed ML packages in R, Python, and Matlab programming languages are inadequate for effective application of classification algorithms and need to be optimized for each new dataset used to generate a model. In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default.

The generalization performance of ML models depends on sample size. The more complex the model, the more data it will need. The dataset we used for our study had 490 samples, however microbiome studies that have smaller sample sizes would benefit from using less complex models such as L2-regularized logistic regression.
